---
title: "Multivariate Analysis -- World Happiness Report"
author: "Manuel Breve, Diego Quintana, Marcel Pons"
date: "2020"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
source(here("R", "utils.R"))

library(psych)
library(tableplot)
library(reshape2)
library(chemometrics)
library(pracma)
library(PerformanceAnalytics)
library(FactoMineR)
library(factoextra)
library(plotrix)
library(gplots)
library(rpart)
library(rpart.plot)
library(randomForest)
library(rsample)

set.seed(42)
```

## Loading Data

```{r}
datasets <- load_whr_data()

# Drop Overall.rank; keep Regional.Indicator for later use
df_row_n <- prepare_numeric_data(datasets$train, drop_rank = TRUE, drop_region = FALSE)
test_row_n <- prepare_numeric_data(datasets$test, drop_rank = TRUE, drop_region = FALSE)

summary(df_row_n)
```

## Imputing Missing Values

One missing value exists in the Perceptions.of.corruption variable.

```{r imputation, include=FALSE}
df_row_n <- impute_missing(df_row_n)
```

```{r rownames, include=FALSE}
df <- column_to_rownames(df_row_n, "Country.or.region")
test <- column_to_rownames(test_row_n, "Country.or.region")
```

## Correlation Analysis

```{r correlation, echo=FALSE}
chart.Correlation(df[, -8], histogram = FALSE)
```

## Assessing Distributions

### Density Histograms

Checking whether all marginals follow a univariate normal distribution.

```{r density, echo=FALSE}
vars <- list(
  list(col = 2, title = "GDP per capita",      y = "Density"),
  list(col = 3, title = "Social Support",      y = ""),
  list(col = 4, title = "Healthy life expect.", y = ""),
  list(col = 5, title = "Choice Freedom",      y = ""),
  list(col = 6, title = "Generosity",          y = "Density"),
  list(col = 7, title = "Perc. Corruption",    y = ""),
  list(col = 1, title = "Score",               y = "")
)

plots <- lapply(vars, function(v) plot_density(df, v$col, v$title, v$y))
grid.arrange(grobs = plots, nrow = 2, ncol = 4)
```

### Q-Q Plots

```{r qqplot, echo=FALSE}
par(mfrow = c(2, 4))
variables <- list(
  list(data = df$GDP.per.capita,               title = "GDP"),
  list(data = df$Social.support,               title = "Social Supp."),
  list(data = df$Healthy.life.expectancy,      title = "Life Exp."),
  list(data = df$Freedom.to.make.life.choices, title = "C Freedom"),
  list(data = df$Generosity,                   title = "Generosity"),
  list(data = df$Perceptions.of.corruption,    title = "P Corruption"),
  list(data = df$Score,                        title = "Ladder Score")
)
for (v in variables) {
  qqnorm(v$data, col = "darkred", main = v$title)
  qqline(v$data, col = "darkblue", lwd = 3)
}
```

### Shapiro-Wilk Tests

```{r shapiro, include=FALSE}
# Only Score passes normality (p-value > 0.05)
shapiro.test(df$GDP.per.capita)
shapiro.test(df$Social.support)
shapiro.test(df$Healthy.life.expectancy)
shapiro.test(df$Freedom.to.make.life.choices)
shapiro.test(df$Generosity)
shapiro.test(df$Perceptions.of.corruption)
shapiro.test(df$Score)
```

### Mahalanobis Distances vs. Chi-Squared Distribution

Verifying whether the multivariate distribution follows a Chi-Squared distribution by simulating 1000 theoretical Chi-Square correlations.

```{r mahalanobis}
X.cov <- var(scale(df[, -8]))
X.mean <- apply(df[, -8], 2, mean)
normal_dist <- mahalanobis(scale(df[, -8]), X.mean, X.cov)

densityplot(normal_dist)
n <- length(normal_dist)
emp_cum <- (c(1:n) - 0.5) / n

plot(
  (mean(normal_dist)) * sort(normal_dist) / (2 * mean(normal_dist)),
  qchisq(emp_cum, df = mean(normal_dist)),
  pch = 20, main = "Q-Q plot Mahalanobis Distances",
  xlab = "Observed", ylab = "Theoretical"
)
aa <- qchisq(emp_cum, df = mean(normal_dist))
bb <- (mean(normal_dist)) * sort(normal_dist) / (2 * mean(normal_dist))
abline(lm(aa ~ bb), col = "blue")

# Simulation: 1000 iterations
(cc.obs <- cor(sort(normal_dist), qchisq(emp_cum, df = 6)))
cc.teo <- NULL
niter <- 1000
for (k in 1:niter) {
  aa <- rchisq(n, df = mean(normal_dist))
  cc.teo <- c(cc.teo, cor(sort(aa), qchisq(emp_cum, df = mean(normal_dist))))
}

# p-value of the observed correlation -- accept: follows Chi-squared
(p.val_cc.obs <- sum(cc.teo < cc.obs) / niter)
```

## Outlier Detection

### Univariate Outliers

```{r boxplots, echo=FALSE}
plots <- lapply(c(2, 3, 4, 5, 6, 7, 1), function(col) plot_boxplot(df, col))
grid.arrange(grobs = plots, nrow = 3, ncol = 3)
```

```{r univariate-outliers}
# Extreme outliers (3x IQR): Denmark, Singapore, Rwanda on Perceptions.of.corruption
(corr.extreme <- detect_univariate_outliers(df, 7, "extreme"))

# Mild outliers (1.5x IQR) across variables
(support.mild  <- detect_univariate_outliers(df, 3, "mild"))
(free.mild     <- detect_univariate_outliers(df, 5, "mild"))
(gene.mild     <- detect_univariate_outliers(df, 6, "mild"))
(corr.mild     <- detect_univariate_outliers(df, 7, "mild"))
```

### Multivariate Outliers

Removing extreme univariate outliers first, then checking Mahalanobis distances.

```{r multivariate-outliers}
remove <- c("Denmark", "Singapore", "Rwanda")
dataset <- df[!row.names(df) %in% remove, ]

outliers <- Moutlier(dataset[, -8], quantile = 0.999, plot = TRUE)

p1 <- densityplot(outliers$md, xlab = "Classical Distances")
p2 <- densityplot(outliers$rd, xlab = "Robust Distances")
grid.arrange(p1, p2, nrow = 2)
```

```{r outlier-check, include=FALSE}
distances <- outliers$md
isOutlier <- (distances >= outliers$cutoff)
table(isOutlier)
```

```{r}
# United Arab Emirates identified as multivariate outlier
out <- data.frame(isOutlier, row.names = row.names(dataset))
subset(out, isOutlier == TRUE)
```

---

## Principal Component Analysis

PCA with the 4 outlier countries as supplementary individuals, Regional.Indicator as supplementary categorical, and Score as supplementary quantitative variable.

```{r pca, include=FALSE}
supplementary_individuals <- which(rownames(df) %in% OUTLIER_COUNTRIES)
res.pca <- PCA(df, quanti.sup = 1, quali.sup = 8, ind.sup = supplementary_individuals,
               scale.unit = TRUE, graph = FALSE)
```

### Individuals Factor Map

```{r pca-individuals, echo=FALSE}
fviz_pca_ind(res.pca,
             habillage = df[-supplementary_individuals, ]$Regional.Indicator,
             col.ind.sup = "black") +
  theme_whr() + xlim(-5.5, 5.5)
```

### Variables Factor Map

```{r pca-variables, echo=FALSE}
fviz_pca_var(res.pca, col.var = "orangered2", repel = TRUE) +
  theme_whr()
```

### Variable Quality and Contributions

```{r pca-quality, echo=FALSE}
vars <- data.frame(res.pca$var$cos2)
vars["Representation"] <- round(vars$Dim.1 + vars$Dim.2, 2)
ggplot(vars, aes(x = rownames(vars), y = Representation, label = Representation)) +
  geom_bar(stat = "identity", fill = "cornflowerblue", color = "black") +
  geom_label() +
  labs(x = "Variable", y = "Cos2 Dim1 + Cos2 Dim2",
       title = "Representation of variables in first factorial plane") +
  theme_whr() + theme(legend.position = "none")

p1 <- fviz_contrib(res.pca, choice = "var", fill = "cornflowerblue",
                   color = "steelblue", axes = 1)
p2 <- fviz_contrib(res.pca, choice = "var", fill = "cornflowerblue",
                   color = "steelblue", axes = 2, ylabel = "")
grid.arrange(p1, p2, nrow = 1)

p1 <- fviz_contrib(res.pca, choice = "ind", fill = "cornflowerblue",
                   color = "steelblue", top = 10, axes = 1)
p2 <- fviz_contrib(res.pca, choice = "ind", fill = "cornflowerblue",
                   color = "steelblue", top = 10, axes = 2)
grid.arrange(p1, p2, nrow = 2)

p1 <- fviz_cos2(res.pca, choice = "ind", fill = "cornflowerblue",
                color = "steelblue", top = 10, axes = 1)
p2 <- fviz_cos2(res.pca, choice = "ind", fill = "cornflowerblue",
                color = "steelblue", top = 10, axes = 2)
grid.arrange(p1, p2, nrow = 2)
```

```{r pca-supplementary, echo=FALSE}
res.pca$ind.sup$cos2
res.pca$quali.sup$v.test
res.pca$quali.sup$cos2
res.pca$quali.sup$eta2
res.pca$quanti.sup$cor
res.pca$quanti.sup$cos2
```

### Biplot

```{r biplot, echo=FALSE}
fviz_pca_biplot(res.pca, geom = c("point", "text"),
                col.var = "orangered4", col.ind.sup = "black",
                col.quanti.sup = "orangered2", title = "Biplot",
                habillage = df[-supplementary_individuals, ]$Regional.Indicator) +
  theme_whr() + xlim(-6.5, 6.5)
```

### Screeplot

```{r screeplot, echo=FALSE}
fviz_screeplot(res.pca, addlabels = TRUE, ggtheme = theme_minimal(),
               barfill = "cornflowerblue", barcolor = "cornflowerblue") +
  theme_whr()
```

---

## Varimax Rotation and Latent Factors

Applying varimax rotation to 5 significant PCA dimensions for interpretable factors.

```{r varimax, echo=FALSE}
significant_dimensions <- 5

coordinates <- data.frame(res.pca$ind$coord[, 1:significant_dimensions])
coordinates_ind.sup <- data.frame(res.pca$ind.sup$coord[, 1:significant_dimensions])
psi <- rbind(coordinates, coordinates_ind.sup)
Phi <- res.pca$var$coord[, 1:significant_dimensions]

pc.rot <- varimax(Phi)
pc.rot$rotmat
pc.rot$loadings
# Factor 1: GDP per capita, Social support, Healthy life expectancy
# Factor 2: Generosity
# Factor 3: Dystopia residual
# Factor 4: Perceptions of corruption
# Factor 5: Freedom to make life choices
```

### Variable Projection (Rotated)

```{r varimax-vars, echo=FALSE}
p_var <- as.data.frame(cbind(pc.rot$loadings[, 1:2]))
angle <- seq(-pi, pi, length = 50)
circle <- data.frame(x = sin(angle), y = cos(angle))

ggplot() +
  theme(aspect.ratio = 1) +
  geom_path(aes(x, y), data = circle, color = "black") +
  geom_text(data = p_var, aes(x = Dim.1, y = Dim.2, label = rownames(p_var))) +
  geom_segment(data = p_var, aes(x = 0, y = 0, xend = Dim.1, yend = Dim.2),
               arrow = arrow(), color = "orangered2") +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(title = "Varimax rotation, variables projection", x = "PC1", y = "PC2") +
  xlim(-1.2, 1) + ylim(-1, 1) +
  coord_fixed() + theme_whr()
```

### Individual Projection (Rotated)

```{r varimax-ind, echo=FALSE}
p_ind <- as.data.frame(cbind(as.matrix(psi) %*% pc.rot$rotmat))
p_ind <- p_ind[, 1:2]
country_region <- df[8]
p_ind <- as.data.frame(merge(p_ind, country_region, by = 0))
colnames(p_ind) <- c("Country", "PC1", "PC2", "Region")

region_colors <- c("black", "lightslateblue", "mediumorchid1", "firebrick1",
                   "green", "lightcoral", "deeppink", "gray55",
                   "darkgoldenrod1", "darkblue")

ggplot(p_ind, aes(x = PC1, y = PC2, label = Country, color = Region)) +
  geom_point() +
  geom_text(check_overlap = FALSE, vjust = -0.5, alpha = 0.5, show.legend = FALSE) +
  scale_colour_manual(values = region_colors) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  labs(title = "Varimax rotation, individuals projection",
       x = "Rotated PC1", y = "Rotated PC2") +
  theme_whr()
```

---

## Hierarchical Clustering (HCPC)

Using Ward's method on the PCA coordinates (outliers excluded) to identify natural country groupings.

```{r hcpc-elbow, echo=FALSE}
hcpc <- HCPC(psi[-c(153:156), ], nb.clust = -1, consol = FALSE, graph = FALSE)
barplot(hcpc$call$t$inert.gain[1:20])
```

```{r hcpc-consolidation, echo=FALSE}
nc <- 5
final_hcpc <- HCPC(psi[-c(153:156), ], nb.clust = nc, graph = FALSE)
```

### Dendrogram

```{r dendrogram}
fviz_dend(final_hcpc, k = 5, cex = 0.7, palette = "jco",
          rect = TRUE, rect_fill = TRUE, rect_border = "jco")
```

### Cluster Characterization

```{r catdes, echo=FALSE}
cut <- final_hcpc$data.clust$clust
table(cut)
catdes(cbind(as.factor(cut), df[-supplementary_individuals, ]), 1, proba = 0.0001)
```

### Cluster Representatives (Parangons)

```{r parangons}
final_hcpc$desc.ind$para
```

### Cluster Visualization

```{r cluster-plot, echo=FALSE}
clusters <- cbind(cut, psi[-c(153:156), ])
clusters$cluster <- paste("Cluster", clusters$cut)
colnames(clusters)[2:6] <- paste0("Dim.", 1:5)

parangons_clusters <- clusters[c(5, 6, 11, 13, 14, 25, 26, 38, 39, 45,
                                 63, 75, 78, 80, 89, 96, 98, 101, 104, 113,
                                 119, 126, 131, 137, 146), ]

region_colors <- c("black", "lightslateblue", "mediumorchid1", "firebrick1",
                   "green", "lightcoral", "deeppink", "gray55",
                   "darkgoldenrod1", "darkblue")

ggplot(clusters, aes(x = Dim.1, y = Dim.2)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_text(aes(label = rownames(clusters), colour = cluster), alpha = 0.5) +
  geom_label(aes(label = rownames(parangons_clusters), colour = cluster),
             alpha = 0.5, data = parangons_clusters) +
  labs(x = "Dim 1", y = "Dim 2",
       title = "Projection of individuals (after consolidation)") +
  scale_colour_manual(values = region_colors) +
  xlim(-7, 7) + theme_whr()
```

---

## Decision Trees

### Building and Pruning

```{r decision-tree}
set.seed(75)
df2 <- df[-supplementary_individuals, -8]

df.tree <- rpart(Score ~ ., data = df2, method = "anova",
                 control = rpart.control(cp = 0.001, xval = 10))
rpart.plot(df.tree, roundint = FALSE)
```

```{r cp-table}
printcp(df.tree)
plotcp(df.tree, minline = TRUE)
```

```{r pruning}
table_tree <- as.data.frame(df.tree$cptable)
min_error_tree <- which.min(table_tree$xerror)
alpha <- table_tree$CP[min_error_tree]
optimum_tree <- prune(df.tree, cp = alpha)
rpart.plot(optimum_tree, roundint = FALSE)
```

### Decision Tree Prediction

```{r dt-prediction}
predicted_dt <- predict(optimum_tree, newdata = test, type = "vector")
evaluate_predictions(test$Score, predicted_dt, "Decision Tree")
```

## Random Forest

### Baseline Model

```{r rf-baseline}
set.seed(75)
rf_model <- randomForest(Score ~ ., data = df2, type = "regression", importance = TRUE)
rf_model

predicted_rf <- predict(rf_model, newdata = test, type = "response")
evaluate_predictions(test$Score, predicted_rf, "Random Forest (baseline)")
```

### Hyperparameter Tuning with tuneRF

```{r rf-tune}
set.seed(42)
res_tuned <- tuneRF(x = df2[, -1], y = df2$Score, ntreeTry = 2000, doBest = TRUE)
print(res_tuned)
plot(res_tuned)

pred_rf_tuned <- predict(res_tuned, newdata = test)
evaluate_predictions(test$Score, pred_rf_tuned, "Random Forest (tuneRF)")
```

### Grid Search

Systematic search over mtry, nodesize, and sampsize to minimize OOB MSE.

```{r rf-gridsearch}
set.seed(42)
mtry_vals <- seq(round(ncol(df2) * 0.2), round(ncol(df2) * 0.8), 1)
nodesize_vals <- seq(3, 8, 2)
sampsize_vals <- round(nrow(df2) * c(0.3, 0.5, 0.7, 0.8))

hyper_grid <- expand.grid(mtry = mtry_vals, nodesize = nodesize_vals,
                          sampsize = sampsize_vals)

mse_err <- matrix(ncol = 6, nrow = nrow(hyper_grid))
for (i in 1:nrow(hyper_grid)) {
  model <- randomForest(
    Score ~ ., data = df2,
    mtry = hyper_grid$mtry[i],
    nodesize = hyper_grid$nodesize[i],
    sampsize = hyper_grid$sampsize[i],
    importance = TRUE
  )
  mse_err[i, ] <- c(mean(model$mse), mean(model$rsq), mean(model$oob.times),
                     hyper_grid$mtry[i], hyper_grid$nodesize[i],
                     hyper_grid$sampsize[i])
}

opt_i <- which.min(mse_err[, 1])
print(cbind(mse_err[opt_i, ]))
```

### Optimal Random Forest

```{r rf-optimal}
rf_optimal <- randomForest(
  Score ~ ., data = df2,
  mtry = hyper_grid$mtry[opt_i],
  nodesize = hyper_grid$nodesize[opt_i],
  sampsize = hyper_grid$sampsize[opt_i],
  importance = TRUE
)
rf_optimal

pred_rf_opt <- predict(rf_optimal, newdata = test)
evaluate_predictions(test$Score, pred_rf_opt, "Random Forest (grid search)")
```

## Final Visualization

```{r final-plot, echo=FALSE}
ggplot(df, aes(x = GDP.per.capita, y = Score,
               size = Generosity, color = Healthy.life.expectancy)) +
  geom_point() +
  labs(title = "GDP vs Happiness Score",
       x = "GDP per capita", y = "Happiness Score",
       size = "Generosity", color = "Life Expectancy") +
  theme_whr()
```
